---
title: "Some example functions to work on"
author: "Joseph Crispell"
date: "23 May 2018"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

-----

## Why write functions?

A function is a stand-a-lone piece of code designed to complete a single task. A function has inputs and usually produces an output, it generally takes the form:

```{r, eval=FALSE}
functionName <- function(input(s)){
  
  # Write R code to a set of tasks here
  
  # Return something from the function - OPTIONAL
  return(output)
}
```

By restricting functions to single tasks, developing code becomes more manageable and efficient. The stand-a-lone aspect of a function is one of its main benefits, it can be developed and tested without relying upon any other code and, once ready, easily incorporated into other functions.

Here we are going to be developing a whole series of functions, to do simple tasks within R. Most of the functions have already been developed and are available directly within R but hopefully by writing them ourselves we'll learn a bit more about the importance of functions and programming in general.

## Functions to work on

### Find the maximum
Write a function that, given a vector of numbers as input, will return the maximum value present in the vector.

### Repeat a value X times
Write a function that, given a value and an integer specifying how many times the value is to be repeated, will return a vector containing the value repeated the specified number of times.

### Calculate the mean
Write a function that, given a vector of numbers as input, will return the sum of all the values in the vector.

### Add two vectors together
Write a function that, given two vectors of numbers ($a$, and $b$) as input, will return a single vector ($output$) where:
\[ output[i] = a[i] + b[i] \]
Although there are much better ways to do this, in your function you should use a `for` loop.

### Check if value is in vector
Write a function that, given a vector and a value as input, checks whether the value is present in the vector

### Get the current date 
Write a function that will return the current date as a string, in the format: dd-mm-YY. You are allowed to use the `Sys.Date()` and `format()` functions.

### Reverse a vector
Write a function that, given an vector of values, will return a vector of the same values but in reverse order.

### Sequence of numbers
Write a function that, given start, end, and step values as input, will return a sequence of numbers starting at the specified start and ending on or before the specified end, incrementing by the specified step value.

### Calculate the euclidean distance between two points
Write a function that, given X and Y coordinates for two points (in the form: $x1$, $y1$, $x2$, $y2$), will return the euclidean (stright line) distance between the points.

### Find the unique values
Write a function that, given a vector as input, will return a vector containing only the unique values present in the input vector.
