---
title: "An introduction to R"
author: "Joseph Crispell"
date: "18 May 2018"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

-----

## What is R?

R is a programming language. A programming language provides a means of communicating instructions to your computer. There many different programming languages out there, for example you may have heard of python, java, or c++. These languages are all very similar but work in slightly different ways making them useful for different tasks.

R was designed for statistics and plotting, but because it is so great it is often used for all sorts of other things! We are going to be using R to create a model that will represent a population infected with a pathogen.

Generally people use RStudio to write programs in R. RStudio is a program that has been designed to make writing R code easy. To get started we'll install R and RStudio on your computers.

## Installation

Download and install R by following the instructions [here](https://www.r-project.org/).

Download and install RStudio by following the instructions [here](https://www.rstudio.com/).

-----

## Getting started with R

Once you got RStudio up and running, let start writing some code:
```{r}
a <- 5
b <- 7
print(paste("a + b =", (a + b)))
```

Although the above R code is simple, it is doing something quite complicated. Firstly, we created two variables `a` and `b`. Variables are used to store data. Here we created a variable called "a" and stored the number `5` in it. Then we created a variable called "b" and stored the number `7` in it.

Variables are important in any programming language, they allow data to be stored and accessible later.

Next we printed some information out into the RStudio console using the `print()` function. A function, in any language, is similar to a variable but instead of storing data, a function stores instructions used to complete a task. The `print()` function will print any information it is provided with to the screen. To find out a bit more about `print()`, type `help(print)` into the RStudio console.

The information we used `print()` to print was a string (set) of characters. We used the function `paste()` function to combine the string "a + b =" with the sum of `a` and `b`.

### Exercise
Create a variable called "c" such that:
\[ c = \frac{a}{b} \]

## Writing a function

We have touched on functions, they store a set of instructions used to complete a task. A function has inputs and usually returns an output. Here is a function that will return the sum of two numbers:
```{r}

addTwoNumbersTogether <- function(firstNumber, secondNumber){
  return(firstNumber + secondNumber)
}

print(paste("a + b =", addTwoNumbersTogether(firstNumber=a, secondNumber=b)))
```

To create a function in R, as in above, you must give it a name and define what you want it to do. We created a function called "addTwoNumbersTogether". The `addTwoNumbersTogether()` function has two inputs: `firstNumber` and `secondNumber`. By naming its inputs, those inputs can be referred to within the function. `addTwoNumbersTogether()` returns the sum of the `firstNumber` and `secondNumber`.

Creating a function will always take this form:
```{r, eval=FALSE}
functionName <- function(input(s)){
  
  # Write R code to a set of tasks here
  
  # Return something from the function - OPTIONAL
  return(output)
}
```

### Exercise
Create a function that returns the product (`a * b`) of two numbers.

## Vectors

What if you have a lot of numbers? You don't want to have to create and name a variable for each one! A `vector` in R can be used to store multiple numbers in a single variable:
```{r}
numbers <- c(2, 3, 23, 13, -1)
```

Now we stored five different numbers in one place. The `c` function in R combines the things within the brackets, `()`, into a vector. To access the numbers within our `numbers` vector, we can either get all of them in one go:
```{r}
print(numbers)
```

Or, we can access each number using its index (position) in the vector. For example to get third number:
```{r}
print(numbers[3])
```

### Exercise
How would I print the first number of the `numbers` vector?

## A `for` loop

A vector provides a means of storing multiple things in the same place. A `for` loop allows you to do the same thing multiple times. For example if I wanted to print each element of our vector `numbers` I could use the following code:
```{r}
print(numbers[1])
print(numbers[2])
print(numbers[3])
print(numbers[4])
print(numbers[5])
```

We managed to print each element in the `numbers` vector, but what if our vector, `numbers`, was larger, it contained more numbers? A better way to print each element of our vector is to use a for loop:
```{r}
for(element in numbers){
  print(element)
}
```

By using the `for` loop, we have printed each element in the `numbers` successfully in a generalisable way that could be applied to any vector.

The `for` loop always takes the form:
```{r,eval=FALSE}
for(element in vector){
  
  # Do something using the element here
}
```

### Exercise
Create a function called "printElements" that will print each element of a vector

## The `if` statement

The last thing we'll touch on is the `if` statement. The `if` statement is programming equivalent of a question, generally taking this form:
```{r, eval=FALSE}
if(question == TRUE){
  # Do something
}else{
  # Do something else
}
```

For example if we wanted to check whether `a` was more than 10, we could use the following code:
```{r}
if(a > 10){
  print("a is more than 10")
}else{
  print("a is not more than 10")
}
```

The `>`, meaning more than, is called an operator. By using operators, such as `>`, `==` (equals), or `<` (less than), we can ask surprisingly complicated questions! We also used the `else` statement, which allows us to do something else, when our condition (`a > 10`) is `FALSE`.

### Exercise
Write code to check whether `b` is more than 3.

## Learning more

In this tutorial we have covered a little about programming in R. There is plenty more to learn, but all of programming more-or-less is built using what we have learnt so far. The important thing is to keep learning, by thinking of tasks that you could use R for and taking advantage of the internet! I have found the following tutorials extremely useful:

* [tutorialspoint](https://www.tutorialspoint.com/r/index.htm)
* [stackoverflow](https://stackoverflow.com/questions/4556524/whats-the-way-to-learn-r)
* [rbloggers](https://www.r-bloggers.com/)

### Exercise
Write a function that will print each even number in a vector of numbers.