---
title: "Simple SIR Model"
author: "Joseph Crispell"
date: "26 January 2018"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Initial parameter settings

The model is a deterministic infection model with three categories: Susceptible ($S$), Infected ($I$), and Recovered ($R$). This SIR model requires two parameters:

* Infectiousness $\beta$
* Recovery $r$

The SIR model can be described with the following equations:
\[
\frac{dS}{dt} = -\frac{\beta * I * S}{N}\\
\frac{dI}{dt} = \frac{\beta * I * S}{N} - r*I\\
\frac{dR}{dt} = r*I\\
\]

Each of the above equations describe the change in the size of a category through time. We'll use the following parameter settings for these equations:
```{r}
# Initial parameter settings
infectiousness <- 0.01
recovery <- 0.1

```

## Building the model

Using the equations described above, an SIR model approximation can be created using the following code:

```{r}

# Set the initial population status

# Initialise a table to store the model output
popStatus <- data.frame(Time=0, S=NA, I=NA, R=NA)

# Run the SIR model


```